Physcraper: A Python package for continually updated gene trees

### 2020-05-18 phylota-overview.Rmd
other similar tools and phylota examples

@@

>>>>

    - @chesters2014protocol present an algorithm to delineate species form GenBank
    DNA data, and cites phylota as a tool that partitions "the contents of a database
    according to homology", by "grouping of database sequences according to internal
    criteria", searching "from a standardized set of references [...] patterns in
    sequence similarity and overlap."
    - the paper presenting phylotaR, a pipeline that recreates the phylota output
    but uses the most updated GenBank release, and is available in R [@bennett2018phylotar],
    cites phylota as its predecessor and inspiration. the authors mention that phylotaR
    pipeline mimics phylota's pipeline but with improvememnts.
    - The paper presenging PhyloBase [@jamil2016visual], cites phylota as one of
    its resources to get phylogenies, along with TreeBASE and others.
    - The paper presenting STBase, a database of one million precomputed species
    trees [@mcmahon2015stbase], cites phylota as a databse of gene trees or multrees,
    "trees having multiple sequences with the same taxon name".
    - @drori2018onetwotree present a Web‐based, user-friendly, online tool for species-tree
    reconstruction, based on the *supermatrix paradigm* and retrieves all available
    sequence data from NCBI GenBank. They cite phylota in the intro as a tool that is "designed to provide
    users with precomputed sets of clusters that were assembled through a single‐linkage
    clustering approach and additionally provides precomputed gene trees that were
    reconstructed for each cluster. In particular, the results obtained by PhyLoTa
    are taxonomically constrained; that is, all sequences of the most recent common
    ancestor are collected even if one specifies only part of a clade".
    - A study developing a tool to link wikipedia data to NCBI taxonomy [@page2011linking]
    cites phylota as a phylogenetic resource that uses the NCBI taxonomy.
    - the study that present DarwinTree [@meng2015darwintree] as well as the study
    presenting an approach to screen sequence data for The Platform
    for Phylogenetic Analysis of Land Plants (PALPP), using the MapReduce paradigm
    to parallelize BLAST [@yong2010screening], both cite phylota as one among other
    "studies based on data mining large numbers of taxa or loci".

@@

    - A phylgeny of 16 fish species of the family Sphyraenidae (Percomorpha), as well
    as two outgroup species of the Centropomidae (barracudas) [@santini2015first]
    - A phylogeny of 34 vole species, Arvicolinae, Rodentia [@garcia2016role]

### 2020-05-18 physcraper_ms.Rmd
other similar tools and phylota examples

- The next step is to identify the search taxon. This must be a taxon (a named clade)
  from the NCBI taxonomy. It will be used to constraint the DNA sequence search on
  the GenBank database within that taxonomic group.
  By default, the search taxon is the most recent common ancestor (MRCA) of the
  matched taxa that is also a named clade in the NCBI taxonomy. This is refered to as the most recent common
  ancestral taxon (MRCAT) or the least inclusive common ancestral taxon (LICA).
  It can be different from the phylogenetic MRCA when the latter is an unnamed clade.
  This is done using the Open Tree API [taxonomy/mrca](https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3#mrca).
  <!-- ***Does physcraper takes into account the focal clade from Open Tree in any way???***
  Not really, it is a search MRCA more than a taxonomic, the default is to get the ingroup from Otol, which is not the focal group.-->
  A search taxon can also be given by the user. It can be a more inclusive
  clade, if the user wants to perform a wider search, outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order. It can also be a less inclusive clade, if the user only
  wants to focus on enriching a particular clade/region within the tree.
  When only one taxon is matched in both the tree and alignment, an MRCAT can be
  found for that single taxon, and thus a DNA sequence search can be performed even
  with only one sequence in the alignment.
- The BLAST algorithm is used to identify similarity among DNA sequences in the GenBank nucleotide
  database within the search taxon and the remaining sequences on the alignment.
  <!--***How does the blast is setup? Is it an all-blast-all??***-->
- The DNA sequence similarity search can be done on a local database that is easily setup by the user. In this case it uses the BLASTn algorithm.
- The search can also be performed remotely, using the bioPython BLASt algorithm.
- A BLAST run is performed for each sequence in the alignment. Results of each BLAST
  are recorded. All matched sequences are saved with their corresponding GenBank
  accesion numbers that will be used to download the whole sequences into a local library.
- Matched sequences below an e-value, percentage similarity, and outside a minimum
  and maximum length threshold are discarded. This will leave out genomic sequences.
  All acepted sequences are asigned an internal identifier, and are further filtered.
- Because we usually do not have the accession number from sequences in the original
  alignment, a filtering process is needed. Accepted sequences that belong to the
  same taxon in the query sequence and that are either identical or shorter than
  the original sequence are also discarded. Only longer sequences belonging to the
  same taxon as the orignal sequence will be considered for further analyses.
- Among the remaining filtered sequences, there are usually several exemplars per taxon.
  Although it can be useful to keep some of them to, for example, investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon kept for further analyses, by default
  5 sequences per taxon are chosen at random. This number can be controlled by the user.
  <!--***How does pyphlawd, or phylota does the exemplar per sepcies choosing?***-->
- Reverse complement sequences are identified and translated.
- This cycle of sequence search is performed two times. ***Is there an arhument to control the number of cycles of blast searches with new sequences***
- A fasta file containing all sequences resulting from the BLAST searches is generated for the user.

@@

@@

@@

>>>>

- The next step is to identify the search taxon. This must be a taxon (a named clade)
  from the NCBI taxonomy. It will be used to constraint the DNA sequence search on
  the GenBank database within that taxonomic group.
  By default, the search taxon is the most recent common ancestor (MRCA) of the
  matched taxa that is also a named clade in the NCBI taxonomy. This is refered to as the most recent common
  ancestral taxon (MRCAT) or the least inclusive common ancestral taxon (LICA).
  It can be different from the phylogenetic MRCA when the latter is an unnamed clade.
  This is done using the Open Tree API [taxonomy/mrca](https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3#mrca).
  <!-- ***Does physcraper takes into account the focal clade from Open Tree in any way???***
  Not really, it is a search MRCA more than a taxonomic, the default is to get the ingroup from Otol, which is not the focal group.-->
  A search taxon can also be given by the user. It can be a more inclusive
  clade, if the user wants to perform a wider search, outside the MRCAT of the matched taxa, e.g., including all taxa within
  the family or the order. It can also be a less inclusive clade, if the user only
  wants to focus on enriching a particular clade/region within the tree.
  When only one taxon is matched in both the tree and alignment, an MRCAT can be
  found for that single taxon, and thus a DNA sequence search can be performed even
  with only one sequence in the alignment.
- The BLAST algorithm is used to identify similarity among DNA sequences in the GenBank nucleotide
  database within the search taxon and the remaining sequences on the alignment.
  <!--***How does the blast is setup? Is it an all-blast-all??***-->
- The DNA sequence similarity search can be done on a local database that is easily
  setup by the user. In this case it uses the BLASTn algorithm.
- The search can also be performed remotely, using the bioPython BLAST algorithm.
- A pairwise all-against-all BLAST search is performed. This means that each sequence
  in the alignment is BLASTed against DNA sequences in the database within the search
  taxon. Results from each one of these BLAST runs are recorded, and matched sequences are saved
  along with their corresponding GenBank accesion numbers. This information will be
  used later to download the whole sequences into a local library.
- Matched sequences below an e-value, percentage similarity, and outside a minimum
  and maximum length threshold are discarded. This filtering leaves out genomic sequences.
  All acepted sequences are asigned an internal identifier, and are further filtered.
- Because the original alignments usually do not have the GenBank accession numbers
  on the sequence names, a filtering process is needed. Accepted sequences that belong to the
  same taxon of the query sequence, and that are either identical or shorter than
  the original sequence are also discarded. Only longer sequences belonging to the
  same taxon as the orignal sequence will be considered for further analyses.
- Among the remaining filtered sequences, there are usually several exemplars per taxon.
  Although it can be useful to keep some of them to, for example, investigate monophyly
  within species, there can be hundreds of exemplar sequences per taxon for some markers.
  To control the number of sequences per taxon kept for further analyses, by default
  5 sequences per taxon are chosen at random. This number can be controlled by the user.
  <!--***How does pyphlawd, or phylota does the exemplar per sepcies choosing?***-->
- Reverse complement sequences are identified and translated.
- This cycle of sequence search is performed two times. ***Is there an argument to control the number of cycles of blast searches with new sequences***
- A fasta file containing all sequences resulting from the BLAST searches is generated for the user.

@@

@chesters2014protocol presents an algorithm that mines GenBank data to delineate species in the insecta.
The authors present a nice comparison with the phylota algorithm.

@@

STBase - @mcmahon2015stbase present a pipeline for species tree construction and
the public database of one million precomputed species trees

@@

OneTwoTree [@drori2018onetwotree] present a Web‐based, user-friendly, online tool for species-tree
reconstruction, based on the *supermatrix paradigm* and retrieves all available
sequence data from NCBI GenBank.
